---
title: "thesis_MinjaeJung_Code"
author: "minjaejung"
date: "2025-11-01"
output: html_document
---
# I. Data Preparation
```{r}
# loading data
library(dplyr)
library(readr)
base_dir <- "c:/temp"
train_df1<-read_csv(file.path(base_dir, 'train_df.csv'))
test_df1<-read_csv(file.path(base_dir, 'test_df.csv'))
train_df1$X<-NULL
test_df1$X<-NULL

# load combined_data for clinical application
combined_data<-read_csv('/temp/combined_df.csv')

###################### transforming for train data ##############################
factor_vars <- c(
    "SEX",
    "DM_hybrid", "HTN_hybrid",
    "ID"
)
# setting variable list( whole variable )
all_vars <- names(train_df1)

# numeric variable list( except for factor variable )
numeric_vars <- setdiff(all_vars, c("ID", factor_vars))

# missing value list
na_values <- c(99999, 77777, 66666)

# 1. transforming the type of variable
for (v in factor_vars) {
    if (v %in% names(train_df1)) {
        train_df1[[v]] <- as.factor(train_df1[[v]])
    }
}
for (v in numeric_vars) {
    if (v %in% names(train_df1)) {
        train_df1[[v]] <- as.numeric(train_df1[[v]])
        train_df1[[v]][train_df1[[v]] %in% na_values] <- NA  # dealing with missing data
    }
}

# transforming for ordinal variable
ordered_levels_list <- list(
    HTN_hybrid = c("Normal", "Prehypertension", "Hypertension"),
    DM_hybrid = c("Normal", "Prediabetes", "Diabetes")
    
)

# repeat
for (var in names(ordered_levels_list)) {
    if (var %in% names(train_df1)) {
        train_df1[[var]] <- factor(
            train_df1[[var]],
            levels = ordered_levels_list[[var]],
            ordered = TRUE
        )
    }
}
summary(train_df1)
str(train_df1)

###################### transforming for test data ##############################
factor_vars <- c(
    "SEX",
    "DM_hybrid", "HTN_hybrid",
    "ID"
)

# setting variable list( whole variable )
all_vars <- names(train_df1)

# numeric variable list( except for factor variable )
numeric_vars <- setdiff(all_vars, c("ID", factor_vars))

# missing value list
na_values <- c(99999, 77777, 66666)


# 1. transforming the type of variable
for (v in factor_vars) {
    if (v %in% names(test_df1)) {
        test_df1[[v]] <- as.factor(test_df1[[v]])
    }
}
for (v in numeric_vars) {
    if (v %in% names(test_df1)) {
        test_df1[[v]] <- as.numeric(test_df1[[v]])
        test_df1[[v]][test_df1[[v]] %in% na_values] <- NA  
    }
}

# transforming for ordinal variable
ordered_levels_list <- list(
    HTN_hybrid = c("Normal", "Prehypertension", "Hypertension"),
    DM_hybrid = c("Normal", "Prediabetes", "Diabetes")
    
)

# repeat
for (var in names(ordered_levels_list)) {
    if (var %in% names(test_df1)) {
        test_df1[[var]] <- factor(
            test_df1[[var]],
            levels = ordered_levels_list[[var]],
            ordered = TRUE
        )
    }
}
summary(test_df1)
str(test_df1)


```

# II. Interaction Term Selection (Horseshoe Prior)
##    Fit horseshoe model (inter_hs)
```{r}
# ------------------------------------------------------------------------------
# Goal: Select significant interaction terms (covariate x TIME) using a
# horseshoe prior, which performs simultaneous shrinkage and variable selection.
#
# All 8 candidate covariates (SEX, AGE, BMI, SYSMEAN, BUN, URICACID,
# HOMOCYST_log, CPEPTIDE) are included as main effects AND interacted with
# TIME (RelativeYear). The horseshoe aggressively shrinks noise interactions
# toward zero while letting genuine signals escape through heavy tails.
#
# Note: `SYSMEAN` in the code corresponds to `SBP` (systolic blood pressure)
# in the manuscript; the two names refer to the same variable.
# ------------------------------------------------------------------------------

library(brms)
# Full model formula with ALL 8 candidate interaction terms.

formm_hs <- bf(
  eGFR ~ SEX + AGE + BMI + SYSMEAN + BUN + URICACID + 
         HOMOCYST_log + CPEPTIDE +
         RelativeYear:SEX +
         RelativeYear:AGE +
         RelativeYear:BMI +
         RelativeYear:SYSMEAN +
         RelativeYear:BUN +
         RelativeYear:URICACID +
         RelativeYear:HOMOCYST_log +
         RelativeYear:CPEPTIDE +
         (1 + RelativeYear | ID)
)
# Prior specification.
# - class="b": horseshoe prior on ALL fixed-effect coefficients (main + interactions).
#   brms default horseshoe() uses df=1 for the local scales lambda_k,
#   equivalent to Half-Cauchy(0,1). The global scale tau also follows
#   Half-Cauchy(0,1).
# - Intercept: weakly informative heavy-tailed t3(0, 10).
# - sd (random-effect SDs): Half-t3(0, 5); avoids the near-zero pathology of
#   Inverse-Gamma priors on variance components (Gelman, 2006).
# - cor: LKJ(eta=2); mild regularization against extreme correlations (|rho| -> 1).
# - sigma (residual SD): Half-t3(0, 10).
prior_hs <- c(
  set_prior("horseshoe()", class = "b"),
  set_prior("student_t(3,0,10)", class = "Intercept"),
  set_prior("student_t(3,0,5)", class = "sd"),
  set_prior("lkj(2)",           class = "cor"),
  set_prior("student_t(3,0,10)", class = "sigma")
)

# Fitting wrapper.
# `save_pars = save_pars(all = TRUE)` is REQUIRED: it retains the local
# shrinkage parameters hs_local[k], which are needed in the next chunk to
# compute the posterior shrinkage factor kappa_k = 1 / (1 + lambda_k^2).
#
# Sampler settings: iter=10000 / warmup=2500 (more than the main model)
# because the funnel geometry of a global-local shrinkage hierarchy needs
# extra warmup. adapt_delta=0.99 reduces divergent transitions.

fit_model <- function(formula, data_input, priors_input) {
    brm(
        formula = formula,
        data = data_input, 
        prior = priors_input,
        family = gaussian(),
        chains = 4, cores = 2,
        iter = 10000, warmup = 2500,
        seed = 123,
        control = list(adapt_delta = 0.99, max_treedepth = 12),
        save_pars = save_pars(all = TRUE)

    )
}
inter_hs<-fit_model(formm_hs, train_df1, prior_hs)
summary(inter_hs)
# saveRDS(inter_hs,'/Users/minjaejung/Desktop/2026/학술지/Revision_0413/inter_hs.rds')
```
## Shrinkage factor & CrI based selection results
```{r}
# ------------------------------------------------------------------------------
# Compute three selection metrics for each candidate interaction:
#   (1) 95% CrI excludes zero (signed agreement of Q2.5 and Q97.5)
#   (2) Posterior mean of the shrinkage factor kappa_k = 1 / (1 + lambda_k^2)
#       Small kappa  ==> data resists global shrinkage ==> genuine signal.
#       kappa near 1 ==> coefficient effectively pulled to zero.
#   (3) Signal-to-noise ratio |E[gamma_k|Y]| / SD(gamma_k|Y)
#
# The paper retains an interaction when all three hold jointly. In this
# dataset, the selected 3 interactions (SEX, URICACID, CPEPTIDE x TIME)
# pass all three criteria, and the excluded 5 fail all three. The final
# filter below uses only conditions (1) and (2); condition (3) is
# automatically satisfied for the retained set.
# ------------------------------------------------------------------------------


library(dplyr)
library(tibble)
library(posterior)

# 1. Fixed-effect summaries (posterior mean, SE, 95% CrI).
fixef_res <- fixef(inter_hs) %>%
  as.data.frame() %>%
  rownames_to_column("Parameter")

# 2. Build an index map between fixed-effect names and hs_local[k].
#    brms stores local scales as hs_local[1], hs_local[2], ... in the order
#    of columns of the design matrix X (excluding the intercept).
standata_info <- make_standata(formm_hs, data = train_df1, prior = prior_hs)
hs_varnames <- colnames(standata_info$X)[-1] # Exclude Intercept

# Keep only interaction terms (names containing ":")
interaction_vars <- grep(":", hs_varnames, value = TRUE)

# 3. Posterior draws
post_draws <- as_draws_df(inter_hs)

# 4. Per-interaction metrics
results_list <- lapply(interaction_vars, function(var_name) {
  # Position of this interaction in the hs_local vector.
  idx <- which(hs_varnames == var_name)
  
  # Local scale draws for this coefficient: lambda_k.
  # (Column naming may differ across brms versions; guard with a check.)
  lambda_col <- paste0("hs_local[", idx, "]")
  
  if (lambda_col %in% colnames(post_draws)) {
    lambda_draws <- post_draws[[lambda_col]]
    # Calculate Shrinkage factor (Kappa): kappa = 1 / (1 + lambda^2)
    # E[kappa_k|Y] summarizes posterior shrinkage toward zero.
    kappa_draws <- 1 / (1 + lambda_draws^2)
    mean_kappa <- mean(kappa_draws)
  } else {
    mean_kappa <- NA
  }
  # Merge with fixed-effect summaries.
  fixed_info <- fixef_res %>% filter(Parameter == var_name)
  
  if(nrow(fixed_info) == 1) {
    estimate <- fixed_info$Estimate
    est_error <- fixed_info$Est.Error
    q2.5 <- fixed_info$Q2.5
    q97.5 <- fixed_info$Q97.5
    # Calculate Signal-to-Noise Ratio (SNR)
    # (SNR) = |E(gamma|Y) / SD(gamma|Y)|
    snr <- abs(estimate / est_error)
    # CrI excludes zero iff its two endpoints share the same sign.
    exclude_zero <- sign(q2.5) == sign(q97.5)
    
    return(tibble(
      Interaction_Term = var_name,
      Estimate = estimate,
      Est_Error = est_error,
      Q2.5 = q2.5,
      Q97.5 = q97.5,
      SNR = snr,
      Mean_Kappa = mean_kappa,
      Exclude_Zero = exclude_zero,
      # `Selected` encodes all three criteria jointly.
      Selected = exclude_zero & (mean_kappa < 0.5) & (snr > 2) 
    ))
  }
})

# 5. Assemble and print.
selection_results <- bind_rows(results_list)

print("=== Horseshoe Prior Interaction Term Selection Results ===")
print(selection_results %>% 
        select(Interaction_Term, Estimate, Q2.5, Q97.5, SNR, Mean_Kappa, Exclude_Zero),
      n = Inf)

# Final selection.
# Note: the filter below uses conditions (1) and (2) only; SNR (condition 3)
# is already satisfied for the retained set (SEX: 15.1, CPEPTIDE: 6.48,
# URICACID: 2.65), so results do not change.
final_selected <- selection_results %>% filter(Exclude_Zero == TRUE & Mean_Kappa < 0.5)
print("=== Final Selected Interaction Terms ===")
print(final_selected$Interaction_Term)
```

# III. Main Model Fitting
```{r}
# ------------------------------------------------------------------------------
# Fit the final BLMM using only the 3 interactions retained from Section II:
#   SEX x TIME, URICACID x TIME, CPEPTIDE x TIME.
# All 8 main effects (including those whose interactions were dropped) remain.
# ------------------------------------------------------------------------------

library(brms)

# Final model formula.
# Note: SYSMEAN here == SBP in the manuscript.
formm<-as.formula(
    paste("eGFR ~  RelativeYear + SEX + AGE + BMI + BUN + URICACID + HOMOCYST_log + CPEPTIDE",
    "+ RelativeYear:CPEPTIDE",
    "+ RelativeYear:URICACID",
    "+ RelativeYear:SEX",
    "+ SYSMEAN + (1 + RelativeYear |ID)"
))


# Weakly informative priors (Model 1 in the prior sensitivity analysis; §V).
# All Student-t_3 / Half-t_3: heavy-tailed, avoids IG pathologies.
priorss <- c(
    set_prior("student_t(3,0,5)", class = "b"),             
    set_prior("student_t(3,0,10)", class = "Intercept"),
    set_prior("student_t(3,0,5)", class = "sd"),
    set_prior("lkj(2)",           class = "cor"),
    set_prior("student_t(3,0,10)", class = "sigma")
)

# Fitting wrapper. iter=8000 / warmup=2000 is sufficient for this well-posed
# model (the funnel geometry of the horseshoe hierarchy is no longer present).
fit_model <- function(formula, data_input, priors_input) {
    brm(
        formula = formula,
        data = data_input, # real train data (eg. train_df)
        prior = priors_input,
        family = gaussian(),
        chains = 4, cores = 2,
        iter = 8000, warmup = 2000,
        seed = 123,
        control = list(adapt_delta = 0.99, max_treedepth = 12),
        save_pars = save_pars(all = TRUE)

    )
}

# Fit the main BLMM.
fit1<-fit_model(formm, train_df1, priorss)
summary(fit1)
# saveRDS(fit1,'/Users/minjaejung/Desktop/2026/학술지/Revision_0413/fit_final.rds')
```

# IV. Prediction & Validation
## Define the function 'predict_with_posterior_update'
```{r}
# ------------------------------------------------------------------------------
# Personalized calibration via single-baseline-eGFR update
# (Kammer et al., 2023; extended to a fully Bayesian posterior update here).
#
# Paper equations (§3.4):
#   r_{i1}    = y_{i1} - x_{i1}^T beta
#   w         = tau_0^2 / (tau_0^2 + sigma^2)
#   E[b_{0i} | y_{i1}] = w * r_{i1}
#   E[b_{1i} | y_{i1}] = (tau_01 / (tau_0^2 + sigma^2)) * r_{i1}
#
# We apply these updates draw-by-draw across all MCMC iterations so that
# parameter uncertainty propagates into the credible intervals of personalized
# predictions -- strictly more honest than plug-in point estimates.
# ------------------------------------------------------------------------------

library(brms)
library(dplyr)
library(purrr)
library(tibble)

# Helper: assign Time_Sequence in {1, 2, 3} in ascending visit order,
# restricted to subjects with exactly 3 visits. (The training/test data are
# constructed to have 3 consecutive waves per subject; this guards against
# accidental violations.)

make_time_seq_rel <- function(dat){
    # 1. Sort by ID and count visits per subject.
    temp_dat <- dat %>%
        group_by(ID) %>%
        arrange(RelativeYear) %>%
        mutate(n_visits = n())
    
    # 2. Defensive filter: drop any ID that does not have exactly 3 visits.
    if (any(temp_dat$n_visits != 3)) {
        warning("Some IDs do not have exactly 3 visits; excluded from evaluation.")
        temp_dat <- temp_dat %>% filter(n_visits == 3)
    }
    
    # 3. Assign sequence indices 1/2/3 to min/median/max RelativeYear.
    temp_dat %>%
        mutate(
            rel_min = first(RelativeYear),
            rel_max = last(RelativeYear),
            # With n_visits == 3 guaranteed, the median visit is the 2nd.
            rel_med = RelativeYear[2], 
            
            Time_Sequence = case_when(
                RelativeYear == rel_min ~ 1,
                RelativeYear == rel_med ~ 2,
                RelativeYear == rel_max ~ 3,
                TRUE ~ NA_real_
            )
        ) %>%
        ungroup() %>%
        filter(!is.na(Time_Sequence))
}

# Main prediction function: returns both pre-update (marginal) and
# post-update (personalized) linear predictors as draws x n matrices.
predict_with_posterior_update <- function(fit, newdata, id_col = "ID",
                                          time_col = "RelativeYear",
                                          outcome_col = "eGFR",
                                          baseline_rule = c("min")) {
    stopifnot(inherits(fit, "brmsfit"))
    baseline_rule <- match.arg(baseline_rule)

    # 1. Baseline row per subject: the earliest RelativeYear (defines t_{i1}=0
    #    in the paper's notation; in the data, RelativeYear is already
    #    centered per subject so min(RelativeYear) == 0).
    base_nd <- newdata %>%
        group_by(.data[[id_col]]) %>%
        slice_min(.data[[time_col]], with_ties = FALSE) %>%
        ungroup()

    # 2. Fixed-effect linear predictors (re_formula = NA drops random effects)
    #    -- draws x n matrices.
    #    lp_fixed_all : at all time points in `newdata`
    #    lp_fixed_base: at the baseline row of each subject
    lp_fixed_all <- brms::posterior_linpred(fit, newdata = newdata, re_formula = NA, transform = FALSE)
    lp_fixed_base <- brms::posterior_linpred(fit, newdata = base_nd, re_formula = NA, transform = FALSE)

    # 3. Extract variance components as draw-level vectors.
    post <- as_draws_df(fit)
    sig_eps <- post[["sigma"]]
    sd0    <- post[["sd_ID__Intercept"]]
    sd1    <- post[["sd_ID__RelativeYear"]]
    
    # Random-effect correlation rho (fallback to 0 if absent).
    if ("cor_ID__Intercept__RelativeYear" %in% names(post)) {
        rho01 <- post[["cor_ID__Intercept__RelativeYear"]]
    } else {
        rho01 <- rep(0, length(sd0))
        message("[info] cor_ID__Intercept__RelativeYear not found → assumed 0 (independent random effects).")
    }

    cov01 <- rho01 * sd0 * sd1
    var0  <- sd0^2
    var_eps <- sig_eps^2

    # 4. Baseline residuals r_{i1}^{(s)} = y_{i1} - x_{i1}^T beta^{(s)}
    #    -- draws x n_subj matrix.
    y0 <- base_nd[[outcome_col]]
    id_levels <- base_nd[[id_col]]
    resid_base <- sweep(lp_fixed_base, 2, y0, FUN = function(hat, y) y - hat)

    # 5. Per-draw posterior update of random effects.
    #    b0_draws = w * resid_base, where w = var0 / (var0 + var_eps)  [paper eq.]
    #    b1_draws = b0_draws * (cov01 / var0)
    #
    #    NOTE on algebraic equivalence:
    #      b1_draws
    #      = b0_draws * (cov01 / var0)
    #      = resid_base * (var0 / (var0 + var_eps)) * (cov01 / var0)
    #      = resid_base * cov01 / (var0 + var_eps)
    #    which matches E[b_{1i}|y_{i1}] = (tau_01 / (tau_0^2 + sigma^2)) * r_{i1}
    #    exactly. The two-step factorization just reuses b0_draws.
    denom <- var0 + var_eps
    b0_draws <- sweep(resid_base, 1, var0 / denom, `*`)
    b1_draws <- sweep(b0_draws,   1, cov01 / var0, `*`)

    # 6. Broadcast per-subject random effects back to per-observation columns.
    nd <- newdata %>%
        mutate(.row_id = row_number()) %>%
        select(.row_id, any_of(c(id_col, time_col, "Time_Sequence")))

    base_map <- base_nd %>%
        mutate(.base_row = row_number()) %>%
        select(all_of(id_col), .base_row)

    nd <- nd %>% left_join(base_map, by = id_col)
    
    # 7. Assemble personalized predictions:
    #    y_tilde^{(s)}_{ij} = x_{ij}^T beta^{(s)}
    #                         + E^{(s)}[b_{0i} | y_{i1}]
    #                         + E^{(s)}[b_{1i} | y_{i1}] * t_{ij}
    draws <- nrow(lp_fixed_all)
    b0_all <- b0_draws[, nd$.base_row, drop = FALSE]
    b1_all <- b1_draws[, nd$.base_row, drop = FALSE]
    time_vec <- matrix(rep(nd[[time_col]], each = draws), nrow = draws)

    lp_updated_all <- lp_fixed_all + b0_all + b1_all * time_vec

    list(
        fixed_linear = lp_fixed_all, # marginal (pre-update) predictions
        updated_linear = lp_updated_all,  # personalized (post-update) predictions
        newdata_index = nd,
        baseline_index = base_nd,
        sigma_draws = sig_eps
    )
}

# ----------------------------------------------------------------------
# Point error metrics (uses posterior mean of predictions as point est.)
# ----------------------------------------------------------------------
point_metrics <- function(y, yhat, p){
    ok <- is.finite(y) & is.finite(yhat)
    y <- y[ok]; yhat <- yhat[ok]
    n <- length(y)
    sse <- sum((y - yhat)^2)
    sst <- sum((y - mean(y))^2)
    r2  <- 1 - sse/sst
    
    # P30 / P10: percentage of predictions within 30% / 10% of the observed
    # value -- commonly reported alongside eGFR prediction studies.
    p30_val <- mean(abs((y - yhat) / y) < 0.3) * 100
    p10_val <- mean(abs((y - yhat) / y) < 0.1) * 100

    tibble(
        RMSE   = sqrt(sse/n),
        MAE    = mean(abs(y - yhat)),
        P30    = p30_val,  # Unit: %
        P10 = p10_val,
        R2     = r2,
        Adj_R2 = 1 - (1 - r2) * (n - 1)/(n - p - 1)
    )
}

# ----------------------------------------------------------------------
# Bayesian R^2 (Gelman, Goodrich, Gabry, Ali, 2019)
#   R2^{(s)} = Var(y_hat^{(s)}) / (Var(y_hat^{(s)}) + sigma^{2(s)})
# Computed per draw; returned as posterior mean and 95% CrI.
#
# Caveat: sigma^{(s)} here is the model's residual SD (population-level),
# not the residual left after personalized updating. Interpret accordingly.
# ----------------------------------------------------------------------
bayesR2_from_draws <- function(yhat_draws, sigma_draws){
    # yhat_draws: draws × n, sigma_draws: draws
    var_yhat <- matrixStats::rowVars(yhat_draws)
    R2_s     <- var_yhat / (var_yhat + sigma_draws^2)
    tibble(
        mean  = mean(R2_s),
        q2.5  = stats::quantile(R2_s, 0.025),
        q97.5 = stats::quantile(R2_s, 0.975)
    )
}


# ----------------------------------------------------------------------
# Combined evaluator: overall + per-time-point point metrics & Bayesian R^2
# for both "Before" (marginal) and "After" (personalized) predictions.
# ----------------------------------------------------------------------
eval_three_timepoints_rel_bayes <- function(fit, test_df){
    require(dplyr)
    require(matrixStats)
    # (a) Restrict to subjects with 3 visits and assign sequence indices.
    dat_ts <- make_time_seq_rel(test_df)
    
    # (b) Run the personalized-calibration predictor.
    pred <- predict_with_posterior_update(fit, dat_ts, id_col = "ID", time_col = "RelativeYear")
    
    # (c) Point predictions = posterior mean across draws.
    mu_before <- colMeans(pred$fixed_linear)    # length of n_all
    mu_after  <- colMeans(pred$updated_linear)
    
    # Number of fixed effects for Adjusted R^2.
    p <- nrow(fixef(fit)) - 1
    
    # observed eGFR
    y <- dat_ts$eGFR
    
    overall_before <- point_metrics(y, mu_before, p) %>% mutate(Time = "All", Type = "Before")
    overall_after  <- point_metrics(y, mu_after,  p) %>% mutate(Time = "All", Type = "After")
    
    # (d) Bayesian R^2, overall (Before / After).
    idx_all <- pred$newdata_index
    split_ids <- split(seq_len(nrow(idx_all)), idx_all$Time_Sequence)
    
    bayesR2_before_overall <- bayesR2_from_draws(pred$fixed_linear,  pred$sigma_draws) %>% mutate(Time="All", Type="Before")
    bayesR2_after_overall  <- bayesR2_from_draws(pred$updated_linear, pred$sigma_draws) %>% mutate(Time="All", Type="After")
    
    # Label map for the three visit indices.
    time_labels <- c("1" = "Baseline", "2" = "Follow-up 1", "3" = "Follow-up 2")
    
    # (e) Per-time-point point metrics.
    by_time_tbl <- purrr::imap_dfr(split_ids, function(cols, k){
        yk <- y[cols]
        mu_b <- mu_before[cols]
        mu_a <- mu_after[cols]
        
        tibble::tibble(
            bind_rows(
                point_metrics(yk, mu_b, p) %>% mutate(Type="Before"),
                point_metrics(yk, mu_a, p) %>% mutate(Type="After")
            )
        ) %>% mutate(Time = time_labels[k])
    })
    
    # Per-time-point Bayesian R^2.
    bayesR2_by_time <- purrr::imap_dfr(split_ids, function(cols, k){
        tibble::tibble(
            bind_rows(
                bayesR2_from_draws(pred$fixed_linear[, cols, drop=FALSE],  pred$sigma_draws) %>% mutate(Type="Before"),
                bayesR2_from_draws(pred$updated_linear[, cols, drop=FALSE], pred$sigma_draws) %>% mutate(Type="After")
            )
        ) %>% mutate(Time = time_labels[k])
    })
    
    list(
        overall_point = bind_rows(overall_before, overall_after),
        overall_bayesR2 = bind_rows(bayesR2_before_overall, bayesR2_after_overall),
        by_time_point = by_time_tbl,
        by_time_bayesR2 = bayesR2_by_time
    )
}
```

## Prediction results
```{r}
# Evaluate the final fit on the held-out test set.
res_final <- eval_three_timepoints_rel_bayes(fit1, test_df1)
print(res_final)
```

## Visualization: predicted vs observed
```{r}
# ------------------------------------------------------------------------------
# Scatterplots of observed vs predicted eGFR at each visit, comparing the
# marginal prediction ("Before") to the personalized update ("After").
# Agreement with the 45-degree line indicates good calibration.
# ------------------------------------------------------------------------------
# Convenience wrapper around posterior_linpred for marginal predictions.

library(tidyverse)
predict_marginal <- function(fit, newdata, id_col, time_col) {
    
    # Fixed-effect linear predictor draws; re_formula = NA drops random effects.
    linpred_fixed <- brms::posterior_linpred(
        fit,
        newdata = newdata,
        re_formula = NA,
        transform = FALSE
    )
    
    # Point estimate = posterior mean.
    pred <- apply(linpred_fixed, 2, mean)
    
    return(list(
        pred = pred,
        linpred_draws = linpred_fixed
    ))
}

# Marginal predictions on the test set.
lp_marg <- brms::posterior_linpred(fit1, newdata = test_df1, re_formula = NA)
test_df1$pred_marg <- apply(lp_marg, 2, mean)

# Personalized (updated) predictions on the test set.
updated_pred <- predict_with_posterior_update(
    fit = fit1,
    newdata = test_df1,
    id_col = "ID",
    time_col = "RelativeYear",
    outcome_col = "eGFR"
)

test_df1$pred_updated <- apply(updated_pred$updated_linear, 2, mean)

# Assign Baseline / Follow-up 1 / Follow-up 2 labels in visit order.
test_df1 <- test_df1 %>%
  group_by(ID) %>%
  arrange(RelativeYear) %>%
  mutate(
    FU = case_when(
      row_number() == 1 ~ "Baseline",
      row_number() == 2 ~ "Follow-up 1",
      row_number() >= 3 ~ "Follow-up 2"
    )
  ) %>%
  ungroup()

test_df1$FU <- factor(
  test_df1$FU,
  levels = c("Baseline", "Follow-up 1", "Follow-up 2")
)

# Long format for ggplot faceting.
df_long <- test_df1 %>%
    select(ID, RelativeYear, eGFR, FU, pred_marg, pred_updated) %>% 
    pivot_longer(
        cols = c(pred_marg, pred_updated),
        names_to = "PredType",
        values_to = "PredValue"
    ) %>%
    mutate(
        PredType = recode(PredType,
                          pred_marg = "Before",
                          pred_updated = "After"
        )
    )

df_long <- df_long %>%
  mutate(
    PredType = factor(
      PredType,
      levels = c("Before", "After")
    )
  )

label_list <- c(
    Before = "Before",
  After  = "After"

)

library(ggplot2)

ggplot(df_long, aes(x = PredValue, y = eGFR, color = FU)) +
    geom_point(alpha = 0.35, size = 1.6) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
    facet_grid(
        PredType ~ FU,
        labeller = labeller(PredType = label_list)
    ) +
    labs(
        x = "Predicted eGFR",
        y = "Observed eGFR"
    ) +
    theme_bw(base_size = 13) +
    theme(
        panel.grid.minor = element_blank(),
        strip.text = element_text(size = 12),
        legend.position = "none"
    )

```


# V. Prior Sensitivity Analysis
## 6 model configurations
```{r}
# ------------------------------------------------------------------------------
# Refit the main model under six prior configurations to check that
# posterior inference is not driven by the specific prior choice.
#
# IMPORTANT TERMINOLOGY:
#   In brms / Stan, `normal(mu, s)` uses s as the **standard deviation**.
#   The manuscript writes "N(0, 75)" with 75 as the **variance**.
#   Hence the code value `normal(0, 8.66)` corresponds to SD = sqrt(75) ~= 8.66,
#   which is the variance-matched counterpart to student_t(3, 0, 5)
#   (whose variance is s^2 * df/(df-2) = 25 * 3 = 75).
#   Similarly, sqrt(300) ~= 17.32 for the t3(0, 10) match.
#
# Why the Normal-intercept prior is centered at 82, not 0:
#   The Student-t_3 has heavy tails and effectively imposes only weak
#   constraints far from 0. A Normal prior with the same variance has
#   Gaussian tails that would be strongly informative against large
#   intercept values. Since the baseline eGFR scale is ~82 mL/min/1.73m^2,
#   we center the Normal intercept prior at 82 so that its mass covers the
#   plausible range for the data-generating process, keeping the comparison
#   with Student-t fair.
#
# IG matching rationale:
#   Var[t3(0, 5)] = 75 and mean(IG(2, 75)) = 75 / (2 - 1) = 75
#   Var[t3(0, 10)] = 300 and mean(IG(2, 300)) = 300 / (2 - 1) = 300
#   so "matched IG" has mean equal to the second moment of the half-t.
# ------------------------------------------------------------------------------

library(brms)

# Same model formula as the main fit.
formm<-as.formula(
    paste("eGFR ~  RelativeYear + SEX + AGE + BMI + BUN + URICACID + HOMOCYST_log + CPEPTIDE",
    "+ RelativeYear:CPEPTIDE",
    "+ RelativeYear:URICACID",
    "+ RelativeYear:SEX",
    "+ SYSMEAN + (1 + RelativeYear |ID)"
))

# Wrapper accepting optional stanvars for custom (Inverse-Gamma) priors that
# brms cannot express directly.
fit_model <- function(formula, data_input, priors_input, stanvars_input = NULL) {
    brm(
        formula = formula,
        data = data_input, 
        prior = priors_input,
        stanvars = stanvars_input,
        family = gaussian(),
        chains = 4, cores = 2,
        iter = 8000, warmup = 2000,
        seed = 123,
        control = list(adapt_delta = 0.99, max_treedepth = 12),
        save_pars = save_pars(all = TRUE)
    )
}


# ==============================================================================
# Prior specifications
# ==============================================================================
# Model 1: Proposed (t3 for coefficients & Half-t3 for SDs).
prior_m1 <- c(
    set_prior("student_t(3, 0, 5)",  class = "b"),
    set_prior("student_t(3, 0, 10)", class = "Intercept"),
    set_prior("student_t(3, 0, 5)",  class = "sd"),
    set_prior("lkj(2)",              class = "cor"),
    set_prior("student_t(3, 0, 10)", class = "sigma")
)
# Model 2: Variance-matched Normal for coefficients, Half-t3 for SDs.
# Intercept centered at 82 (baseline eGFR scale) -- see header note.
prior_m2 <- c(
    set_prior("normal(0, 8.66)",    class = "b"),
    set_prior("normal(82, 17.32)",  class = "Intercept"), # 수정됨
    set_prior("student_t(3, 0, 5)", class = "sd"),
    set_prior("lkj(2)",             class = "cor"),
    set_prior("student_t(3, 0, 10)",class = "sigma")
)


# Model 3: t3 for coefficients, matched Inverse-Gamma for variances.
# IG priors are applied on sigma^2 and sd_1^2 via stanvars (below).
# Empty priors for sd/sigma tell brms to skip its default and use stanvars.
prior_m3_base <- c(
    set_prior("student_t(3, 0, 5)",  class = "b"),
    set_prior("student_t(3, 0, 10)", class = "Intercept"),
    set_prior("lkj(2)",              class = "cor"),
    set_prior("", class = "sd"),    
    set_prior("", class = "sigma")  
)

# Matched IG on the variance scale. `log(2 * sigma)` and `log(2 * sd_1[n])`
# are Jacobian adjustments for the parameterization change from variance to
# SD (brms parameterizes in SD; we declare an IG on variance).
stanvars_matched_IG <- stanvar(scode = "
  target += inv_gamma_lpdf(square(sigma) | 2, 300) + log(2 * sigma);
  for (n in 1:M_1) {
    target += inv_gamma_lpdf(square(sd_1[n]) | 2, 75) + log(2 * sd_1[n]);
  }
", block = "model")

# Model 4: matched Normal for coefficients & matched IG for variances.
prior_m4_base <- c(
    set_prior("normal(0, 8.66)",    class = "b"),
    set_prior("normal(82, 17.32)",  class = "Intercept"), # 수정됨
    set_prior("lkj(2)",             class = "cor"),
    set_prior("", class = "sd"),    
    set_prior("", class = "sigma")  
)

# Model 5: t3 for coefficients & non-informative IG(0.01, 0.01) for variances.
prior_m5_base <- prior_m3_base

stanvars_noninfo_IG <- stanvar(scode = "
  target += inv_gamma_lpdf(square(sigma) | 0.01, 0.01) + log(2 * sigma);
  for (n in 1:M_1) {
    target += inv_gamma_lpdf(square(sd_1[n]) | 0.01, 0.01) + log(2 * sd_1[n]);
  }
", block = "model")

# Model 6: t3 & Weakly Informative Inverse-Gamma IG(1, 1)
prior_m6_base <- prior_m3_base

stanvars_mild_IG <- stanvar(scode = "
  target += inv_gamma_lpdf(square(sigma) | 1, 1) + log(2 * sigma);
  for (n in 1:M_1) {
    target += inv_gamma_lpdf(square(sd_1[n]) | 1, 1) + log(2 * sd_1[n]);
  }
", block = "model")


# ==============================================================================
# Fit and cache all six models
# ==============================================================================
# Replace `save_dir` with a local path before running.
save_dir <- "/Users/minjaejung/Desktop/2026/학술지/Revision_0413"
dir.create(save_dir, showWarnings = FALSE)

cat("Fitting Model 1...\n")
fit_m1 <- fit_model(formm, train_df1, prior_m1)
saveRDS(fit_m1, file.path(save_dir, "Model1.rds"))

cat("Fitting Model 2...\n")
fit_m2 <- fit_model(formm, train_df1, prior_m2)
saveRDS(fit_m2, file.path(save_dir, "Model2.rds"))

cat("Fitting Model 3...\n")
fit_m3 <- fit_model(formm, train_df1, prior_m3_base, stanvars_input = stanvars_matched_IG)
saveRDS(fit_m3, file.path(save_dir, "Model3.rds"))

cat("Fitting Model 4...\n")
fit_m4 <- fit_model(formm, train_df1, prior_m4_base, stanvars_input = stanvars_matched_IG)
saveRDS(fit_m4, file.path(save_dir, "Model4.rds"))

cat("Fitting Model 5...\n")
fit_m5 <- fit_model(formm, train_df1, prior_m5_base, stanvars_input = stanvars_noninfo_IG)
saveRDS(fit_m5, file.path(save_dir, "Model5.rds"))

cat("Fitting Model 6...\n")
fit_m6 <- fit_model(formm, train_df1, prior_m6_base, stanvars_input = stanvars_mild_IG)
saveRDS(fit_m6, file.path(save_dir, "Model6.rds"))

cat("All 6 models have been successfully fitted and saved to:", save_dir, "\n")

```


##    Posterior density comparison plots
```{r}
# ------------------------------------------------------------------------------
# Overlay marginal posterior densities of each fixed-effect parameter across
# the six prior configurations. If the curves coincide, posterior inference
# is robust to the prior choice.
# ------------------------------------------------------------------------------

library(brms)
library(posterior)
library(tidyverse)
library(patchwork)

# 1. Collect all fitted models.
fit_list <- list(
    "Model 1" = fit_m1,
    "Model 2" = fit_m2,
    "Model 3" = fit_m3,
    "Model 4" = fit_m4,
    "Model 5" = fit_m5,
    "Model 6" = fit_m6
)

# 2. Target fixed-effect parameter names (brms uses the 'b_' prefix).
target_vars <- c(
    "b_Intercept", "b_RelativeYear", "b_SEXMale", "b_AGE", "b_BMI", 
    "b_BUN", "b_URICACID", "b_HOMOCYST_log", "b_CPEPTIDE", "b_SYSMEAN", 
    "b_RelativeYear:CPEPTIDE", "b_RelativeYear:URICACID", "b_RelativeYear:SEXMale"
)

# 3. Extract draws from each model, tagging with a Model column.
draws_list <- lapply(names(fit_list), function(m_name) {
    fit <- fit_list[[m_name]]
    
    # Extract draws using as_draws_df(), select target variables, and add a Model identifier column
    draws <- as_draws_df(fit) %>%
        as_tibble() %>%
        select(any_of(target_vars)) %>%
        mutate(Model = m_name)
    
    return(draws)
})

# Bind the list of dataframes into one
all_draws <- bind_rows(draws_list)

# 4. Wide -> long, and strip the 'b_' prefix for nicer axis labels.
draws_long <- all_draws %>%
    pivot_longer(cols = starts_with("b_"), names_to = "Parameter", values_to = "Value") %>%
    mutate(Parameter = str_remove(Parameter, "^b_"))

params_to_plot <- c(
    "Intercept", "RelativeYear", "SEXMale", "AGE", "BMI", "BUN", 
    "URICACID", "HOMOCYST_log", "CPEPTIDE", "SYSMEAN", 
    "RelativeYear:CPEPTIDE", "RelativeYear:URICACID", "RelativeYear:SEXMale"
)

# 5. For each parameter: two side-by-side density plots.
#    Left  panel = Models 1,2,3,4 (varies coefficient prior; matched variance prior)
#    Right panel = Models 1,5,6   (varies variance prior; fixed coefficient prior)
for (param in params_to_plot) {
    
    # Filter data for the current parameter
    param_data <- draws_long %>% filter(Parameter == param)
    
    if(nrow(param_data) == 0) {
        message(paste("Skipping", param, "- no data found."))
        next
    }
    
    # Left: effect of different coefficient priors.
    plot1 <- param_data %>%
        filter(Model %in% c("Model 1", "Model 2", "Model 3", "Model 4")) %>%
        ggplot(aes(x = Value, fill = Model, color = Model)) +
        geom_density(alpha = 0.4, linewidth = 0.8) +
        labs(title = "Models 1, 2, 3, 4", x = "Estimate (Posterior)", y = "Density") +
        theme_minimal() +
        scale_fill_viridis_d() + scale_color_viridis_d() +
        theme(legend.position = "bottom")
    
    # Right: effect of different Inverse-Gamma specifications on variances.
    plot2 <- param_data %>%
        filter(Model %in% c("Model 1", "Model 5", "Model 6")) %>%
        ggplot(aes(x = Value, fill = Model, color = Model)) +
        geom_density(alpha = 0.4, linewidth = 0.8) +
        labs(title = "Models 1, 5, 6", x = "Estimate (Posterior)", y = "") +
        theme_minimal() +
        scale_fill_brewer(palette = "Set1") + scale_color_brewer(palette = "Set1") +
        theme(legend.position = "bottom")
    
    # --- Combine plots using the patchwork package ---
    combined_plot <- plot1 + plot2 + 
        plot_annotation(
            title = paste("Prior Sensitivity Analysis:", param),
            subtitle = "Comparison of Posterior Distributions across models"
        )
    
    # Display the plot in the R environment
    print(combined_plot)
    
    # Save as PNG. Replace ':' in names to keep filenames filesystem-safe.
    safe_name <- gsub(":", "_", param)
    file_name <- paste0("Posterior_Density_", safe_name, ".png")
    
    ggsave(filename = file_name, plot = combined_plot, 
           width = 12, height = 6, dpi = 300)
}

message("Posterior density visualization and saving completed for all parameters.")
```


##    Predictive performance across configurations
```{r}
# Re-run the evaluator on the held-out test set for each of the 6 models.
# If all six yield nearly identical metrics, posterior predictive inference
# is robust to the prior configuration.
res1 <- eval_three_timepoints_rel_bayes(fit_m1, test_df1)
print(res1)
res2 <- eval_three_timepoints_rel_bayes(fit_m2, test_df1)
print(res2)
res3 <- eval_three_timepoints_rel_bayes(fit_m3, test_df1)
print(res3)
res4 <- eval_three_timepoints_rel_bayes(fit_m4, test_df1)
print(res4)
res5 <- eval_three_timepoints_rel_bayes(fit_m5, test_df1)
print(res5)
res6 <- eval_three_timepoints_rel_bayes(fit_m6, test_df1)
print(res6)
```


# VI. Clinical Application (Rapid Decliner Identification)
```{r}
# ------------------------------------------------------------------------------
# Prospective validation: can the BLMM, fitted to waves 4-6, identify
# individuals who become rapid decliners in later waves?
#
# Observed slope definition:
#   obs_slope = (eGFR_last - eGFR_baseline) / (t_last - t_baseline)
# where t_baseline = wave 4 and t_last is the subject's latest observation
# among waves 7 and 8. Waves 9-10 are EXCLUDED because the creatinine assay
# and reagent were changed starting from wave 9, breaking comparability with
# the training data. Waves 5-6 are excluded because they are already in the
# training set.
#
# Predictor:
#   For each subject i, posterior draws of the individual slope
#       slope_i^{(s)} = beta_1^{(s)} + b_{1i}^{(s)} + Z_i^T gamma^{(s)}
#   are used to compute Pr(slope_i < c | Y) for c in {-2, -3}. This
#   probability is then used as a continuous predictor in an ROC analysis
#   against the observed-slope indicator.
#
# Prerequisites (objects expected to be in the workspace):
#   - train_df1, test_df1  : training / test data
#   - fit1                 : main BLMM fit (from Section III)
#   - combined_data        : long-term eGFR follow-up (waves 7-8)
# ------------------------------------------------------------------------------
```

## Long-term slope estimation
```{r}


library(dplyr)
library(purrr)
library(brms)
library(tibble)


# ----------------------------------------------------------------------
# (1) Build the prediction dataset:
#     - Baseline covariates taken from train_df (wave 4 row per subject).
#     - Long-term eGFR observations from combined_data (waves 7-8).
#     - "eGFR" column of newdata_pred is set to the BASELINE eGFR so that
#       predict_with_posterior_update() uses it for the single-obs update.
# ----------------------------------------------------------------------

make_newdata_for_prediction <- function(train_df, combined_data){
    
    # Baseline covariates + baseline eGFR, renamed to avoid collisions.
    base_cov <- train_df %>%
        group_by(ID) %>%
        slice_min(RelativeYear, with_ties = FALSE) %>%
        ungroup()
    
    base_cov <- base_cov %>% 
        rename(eGFR_baseline = eGFR,
               RelativeYear_baseline = RelativeYear)
    
    # Keep only IDs that appear in BOTH training and long-term follow-up.
    valid_ids <- intersect(unique(train_df$ID), unique(combined_data$ID))
    
    combined_sub <- combined_data %>%
        filter(ID %in% valid_ids)
    
    # Join baseline covariates onto every long-term row.
    newdata <- combined_sub %>%
        left_join(base_cov, by = "ID") %>%
        arrange(ID, RelativeYear)
    
    # Ground-truth long-term eGFR for evaluation.
    observed <- newdata %>%
        select(ID, RelativeYear, eGFR)
    
   # Predictor frame: overwrite eGFR with baseline eGFR so it is the
   # single-observation calibration input. Time_Sequence = 1 flags this
   # baseline row for internal code that relies on the label.
    newdata_pred <- newdata %>%
        rename(eGFR_longterm = eGFR) %>%
        select(-eGFR_longterm) %>%        
        mutate(eGFR = eGFR_baseline,       
               RelativeYear = RelativeYear,
               Time_Sequence = 1    
        )
    
    return(list(
        observed = observed,
        newdata_pred = newdata_pred
    ))
}

# -------------------------------------------------------------
# (2) Personalized prediction at long-term time points.
# -------------------------------------------------------------
predict_longterm_bayes <- function(fit, train_df, combined_data){
    
    nd <- make_newdata_for_prediction(train_df, combined_data)
    newdata_pred <- nd$newdata_pred
    observed      <- nd$observed
    
    # Posterior-updated prediction (reuses Section IV machinery).
    pred <- predict_with_posterior_update(
        fit,
        newdata = newdata_pred,
        id_col = "ID",
        time_col = "RelativeYear"
    )
    
    # Point estimate = posterior mean.
    mu_after <- colMeans(pred$updated_linear)
    
    # Attach observed values for downstream comparison.
    out <- newdata_pred %>%
        mutate(pred_eGFR = mu_after) %>%
        left_join(observed, by = c("ID","RelativeYear")) %>%
        rename(obs_eGFR = eGFR.y)
    
    return(out)
}

longterm_result <- predict_longterm_bayes(
    fit = fit1,
    train_df = train_df1,
    combined_data = combined_data
)


# ----------------------------------------------------------------------
# (3) Build posterior draws of each subject's individual slope:
#         slope_i^{(s)} = beta_1^{(s)} + b_{1i}^{(s)} + Z_i^T gamma^{(s)}
#     We then compute Pr(slope_i < c | Y) by Monte Carlo averaging.
# ----------------------------------------------------------------------
library(brms)
library(dplyr)
library(purrr)
library(tibble)

# Posterior draws from the main fit.
post <- as_draws_df(fit1)

# Population slope(β₁) (length = n_draws).
b_rel <- post[["b_RelativeYear"]]

# Interaction coefficients gamma (length = n_draws each).
b_sex_time <- post[["b_RelativeYear:SEXMale"]] 
b_uric_time <- post[["b_RelativeYear:URICACID"]]
b_cpep_time <- post[["b_RelativeYear:CPEPTIDE"]]

# Random slope matrix r_ID[i, RelativeYear]: draws x n_subj.
sl_cols <- grep("^r_ID\\[.*,RelativeYear\\]$", names(post), value = TRUE)
rand_sl_mat <- as.matrix(post[, sl_cols])

# Recover subject IDs from column names in brms' r_ID[..] format.
id_vec <- sl_cols |>
    sub("^r_ID\\[", "", x = _) |>
    sub(",RelativeYear\\]$", "", x = _)

# Align baseline covariates to id_vec order.
base_cov_by_id <- train_df1 %>%
    group_by(ID) %>%
    slice_min(RelativeYear, with_ties = FALSE) %>%
    ungroup() %>%
    mutate(ID = as.character(ID)) %>%
    filter(ID %in% id_vec) %>%
    arrange(match(ID, id_vec))

# SEX dummy (Female = 0, Male = 1). URICACID and CPEPTIDE are already
# z-scored in train_df1 (done during preprocessing).
sex_male <- as.integer(base_cov_by_id$SEX == "Male")
uric_vec <- base_cov_by_id$URICACID 
cpep_vec <- base_cov_by_id$CPEPTIDE

# Assemble per-subject slope draws: (draws x n_subj).
# fixed_slope       = beta_1^{(s)} + b_{1i}^{(s)}  (broadcast beta_1 across subjects)
# interaction_contrib = sum over k of gamma_k^{(s)} * Z_{ik}
fixed_slope <- sweep(rand_sl_mat, 1, b_rel, FUN = "+")

interaction_contrib <- outer(b_sex_time, sex_male) +
                       outer(b_uric_time, uric_vec) +
                       outer(b_cpep_time, cpep_vec)

slope_draws <- fixed_slope + interaction_contrib


id_vec <- sl_cols |>
    sub("^r_ID\\[", "", x = _) |>
    sub(",RelativeYear\\]$", "", x = _)

# Per-subject posterior summary and exceedance probabilities.
slope_posterior_summary <- map_dfr(
    seq_along(id_vec),
    function(j){
        s <- slope_draws[, j]
        tibble(
            ID          = id_vec[j],
            slope_mean  = mean(s),
            slope_median= median(s),
            sd_slope    = sd(s),
            pr_lt_2     = mean(s < -2),   # Pr(slope < -2)
            pr_lt_3     = mean(s < -3),   # Pr(slope < -3)
            pr_lt_5     = mean(s < -5)    # Pr(slope < -5) KDIGO rapid progression
        )
    }
)

# Optional posterior-probability flag (prob >= 0.80).
slope_posterior_summary <- slope_posterior_summary %>%
    mutate(
        rapid2_80 = pr_lt_2 >= 0.80,
        rapid3_80 = pr_lt_3 >= 0.80,
        rapid5_80 = pr_lt_5 >= 0.80
    )

slope_posterior_summary


# ----------------------------------------------------------------------
# (4) Observed slope = simple two-point linear rate between baseline
#     (wave 4) and the latest long-term observation (wave 7 or 8).
# ----------------------------------------------------------------------
obs_slope_df <- longterm_result %>%
    group_by(ID) %>%
    summarise(
        time0      = first(RelativeYear_baseline),
        eGFR0      = first(eGFR_baseline),
        time_last  = max(RelativeYear),
        eGFR_last  = obs_eGFR[which.max(RelativeYear)],
        obs_slope  = (eGFR_last - eGFR0) / (time_last - time0)
    ) %>%
    ungroup()
# Merge predicted probabilities with observed rapid-decliner labels.
slope_compare <- slope_posterior_summary %>%
    mutate(ID = as.character(ID)) %>%  
    inner_join(
        obs_slope_df %>% mutate(ID = as.character(ID)), 
        by = "ID"
    ) %>%
    mutate(
        rapid2_true = obs_slope <= -2,   # common epidemiological threshold
        rapid3_true = obs_slope <= -3,   # common clinical threshold
        rapid5_true = obs_slope <= -5    # KDIGO; effectively zero events in this cohort
    )

slope_compare

library(dplyr)
library(ggplot2)
library(pROC)

# Reference distribution of true labels (for context):
# > table(slope_compare$rapid2_true)
# 
# FALSE  TRUE 
#  1380   123 
# > table(slope_compare$rapid3_true)
# 
# FALSE  TRUE 
#  1477    26 
# > table(slope_compare$rapid5_true)
# 
# FALSE  TRUE 
#  1502     1 

# NOTE: 5-unit threshold is too rare for a stable ROC; we report only
# the -3 and -2 analyses below.

# ROC analysis for the -3 threshold.
dat_roc3 <- slope_compare %>%
    filter(!is.na(pr_lt_3), !is.na(rapid3_true))

roc3 <- roc(
    response  = dat_roc3$rapid3_true,
    predictor = dat_roc3$pr_lt_3,
    quiet     = TRUE
)

auc3 <- auc(roc3)
auc3
# (reference) Area under the curve: ~0.809

dat_roc2 <- slope_compare %>%
    filter(!is.na(pr_lt_2), !is.na(rapid2_true))

roc2 <- roc(
    response  = dat_roc2$rapid2_true,
    predictor = dat_roc2$pr_lt_2,
    quiet     = TRUE
)

auc2 <- auc(roc2)
auc2
# (reference) Area under the curve: ~0.736
```


## ROC / AUC, Sensitivity, Specificity, NPV, PPV
```{r}
# ------------------------------------------------------------------------------
# Extract the full set of classification metrics at the Youden-optimal
# threshold (maximizes Sensitivity + Specificity - 1).
#
# Clinical note: in a low-prevalence setting (rapid decliners are 1.7% at
# threshold -3 and 8.2% at -2), PPV is structurally low even when AUC and
# NPV are strong. The model's useful role here is NPV-based: reliable
# exclusion of rapid decline rather than positive identification.
# ------------------------------------------------------------------------------
library(pROC)
library(caret)
library(dplyr)

# ============================================================
# Function: compute metrics at the Youden-optimal threshold.
# ============================================================
get_performance_metrics <- function(response_vec, predictor_vec, target_label) {
    
    # 1. Build ROC object and compute AUC.
    roc_obj <- roc(response = response_vec, predictor = predictor_vec, quiet = TRUE)
    auc_val <- as.numeric(auc(roc_obj))
    
    # 2. Youden-optimal threshold. pROC's "best" returns the point(s)
    #    maximizing sensitivity + specificity.
    coords_best <- coords(roc_obj, "best", ret = c("threshold", "sensitivity", "specificity"), 
                          transpose = FALSE)
    
    # If multiple optima, take the first.
    best_thresh <- coords_best$threshold[1]
    
    # 3. Hard classification: predictor >= threshold ==> Positive.
    pred_class <- factor(ifelse(predictor_vec >= best_thresh, TRUE, FALSE), levels = c("TRUE", "FALSE"))
    true_class <- factor(response_vec, levels = c("TRUE", "FALSE"))
    
    # 4. Confusion matrix and summary metrics.
    cm <- confusionMatrix(pred_class, true_class, positive = "TRUE")
    
    # 5. Collect into a named vector for compact reporting.
    metrics <- c(
        Target          = target_label,
        AUC             = round(auc_val, 3),
        Best_Threshold  = round(best_thresh, 3),
        Sensitivity     = round(cm$byClass["Sensitivity"], 3),
        Specificity     = round(cm$byClass["Specificity"], 3),
        PPV             = round(cm$byClass["Pos Pred Value"], 3), # Precision
        NPV             = round(cm$byClass["Neg Pred Value"], 3),
        F1_Score        = round(cm$byClass["F1"], 3),
        Accuracy        = round(cm$overall["Accuracy"], 3)
    )
    
    return(metrics)
}

# ============================================================
# Compute metrics for both rapid-decliner definitions.
# ============================================================

# Threshold -3 (standard clinical cutoff, rare events in this cohort).
perf_rapid3 <- get_performance_metrics(
    response_vec  = dat_roc3$rapid3_true,
    predictor_vec = dat_roc3$pr_lt_3,
    target_label  = "Rapid Decliner (<-3)"
)

# Threshold -2 (more sensitive definition, more events).
perf_rapid2 <- get_performance_metrics(
    response_vec  = dat_roc2$rapid2_true,
    predictor_vec = dat_roc2$pr_lt_2,
    target_label  = "Rapid Decliner (<-2)"
)

# ============================================================
# Combine & Print Result Table
# ============================================================
final_perf_table <- rbind(perf_rapid3, perf_rapid2) %>% 
    as.data.frame() %>%
    remove_rownames() %>%
    column_to_rownames("Target")

print(final_perf_table)
# Expected output (reference values):
#                        AUC Best_Threshold Sensitivity Specificity    PPV    NPV F1_Score Accuracy
# Rapid Decliner (<-3) 0.809          0.009       0.885       0.609  0.038  0.997    0.073    0.614
# Rapid Decliner (<-2) 0.736          0.125       0.780       0.566  0.138  0.967    0.235    0.583

# Optional: write to CSV for inclusion in the manuscript.
# write.csv(final_perf_table, "Table3_Performance_Metrics.csv")
```
# End of analysis script.

